# Exercise 1 {#ex1}

## Load packages

Here is the R code to download the required packages for this exercise.

```{r pac-load, eval = TRUE}
# install package manager 'pacman'
if (!require(pacman)){
  install.packages('pacman')
}

# load packages needed for this exercise
pacman::p_load(tidyverse, # for data manipulation
							 psych, # for `describe()` function
							 ggpubr, # for graphing
							 rlang, # for tidy evaluation
							 ggfortify)
```

## Data

This is equivalent to the data step in SAS. Here, the data is imported from a file `ex1.csv` using the function `read_csv`. This function will download the data file direclty from [here](https://raw.githubusercontent.com/luckymehra/epidem-exercises/master/data/ex1.csv).

```{r data-import, eval=TRUE}
# Import data
a <- read_csv("data/ex1.csv") %>% 
	# create a new variable `DIFF`
	mutate(DIFF = ESTIMATE - ACTUAL)

# print the data
a
```

## Descriptive statistics

```{r desc-stat, eval=TRUE}
# use the describe() function to get an idea about different variables
sm_a <- a %>% 
	group_by(TEST) %>% 
	summarise_at(vars(DIFF),
							 list(~mean(., na.rm = TRUE),
							 		 ~mode(.),
							 		 ~median(., na.rm = TRUE),
							 		 ~sd(., na.rm = TRUE),
							 		 ~var(., na.rm = TRUE),
							 		 ~IQR(., na.rm = TRUE),
 							 		 ~min(., na.rm = TRUE),
							 		 ~max(., na.rm = TRUE),
							 		 ~n(),
							 		 ~skew(.),
							 		 ~kurtosi(.)
							 		 )) %>% 
	mutate(CV = (sd/mean)*100,
				 std.err = sd/sqrt(n)) %>% 
	ungroup()

sm_a
```

## Create histogram, boxplot, and qqplot

```{r plots, eval=TRUE}
# histograms
create_hist <- function(test_id){
a %>%
	filter(TEST == {{test_id}}) %>% 
	gghistogram("DIFF", bins = 6, fill = "skyblue") +
		coord_flip()
}

create_hist(unique(a$TEST)[1]) # BASELINE
create_hist(unique(a$TEST)[2]) # POST1
create_hist(unique(a$TEST)[3]) # POST2

# boxplots
create_box <- function(test_id){
	a %>% 
		filter(TEST == {{test_id}}) %>% 
		ggboxplot(y = "DIFF", fill =  "skyblue",
							main = {{test_id}})
}

create_box(unique(a$TEST)[1]) # BASELINE
create_box(unique(a$TEST)[2]) # POST1
create_box(unique(a$TEST)[3]) # POST2

# distribution of DIFF by TEST
ggboxplot(data = a,
					x = "TEST", y = "DIFF")
# qqplots
create_qq <- function(test_id){
a %>%
	filter(TEST == {{test_id}}) %>% 
	ggqqplot("DIFF",
					 main = {{test_id}}) 
}

create_qq(unique(a$TEST)[1]) # BASELINE
create_qq(unique(a$TEST)[2]) # POST1
create_qq(unique(a$TEST)[3]) # POST2
```

## Normality test

```{r norm-test, eval = TRUE}
st <- shapiro.test(a$DIFF)
data.frame(st$method, st$statistic, st$p.value)

do_shp_test <- function(df){
	st <- shapiro.test(df$DIFF)
	data.frame(st$method, st$statistic, st$p.value)
}

a %>% 
	group_by(TEST) %>% 
	do(do_shp_test(.)) %>% 
	ungroup()
```

## Linear regression

```{r lin-reg}
# fit linear regression by TEST
# Response: ESTIMATE
est_lm <- function(df){
	model <- lm(ESTIMATE ~ ACTUAL, data = df)
	
	# fit plot
	fit_plot <- ggplot(data = df, aes(x = ACTUAL, y = ESTIMATE)) +
		geom_point()+
		stat_smooth(method = "lm", col = "blue")
	
	return(list(summary(model), autoplot(model), fit_plot))
}

by(a, a$TEST, est_lm)

# Response: DIFF
diff_lm <- function(df){
	model <- lm(DIFF ~ ACTUAL, data = df)
	
	# fit plot 
	fit_plot <- ggplot(data = df, aes(x = ACTUAL, y = DIFF)) +
		geom_point()+
		stat_smooth(method = "lm", col = "blue")
		
	return(list(summary(model), autoplot(model)))
}

by(a, a$TEST, diff_lm)


```