# Exercise 4 {#ex4}



## Load packages

Here is the R code to download the required packages for this exercise.

```{r pac-load}
# install package manager 'pacman'
if (!require(pacman)){
  install.packages('pacman')
}
```

Load the packages needed for this exercise:

```{r}
pacman::p_load(tidyverse, 
       nlme, 
       emmeans)
```

## Import data

Our data is located in `ex4.csv` file, which can be found on [my github repo](https://raw.githubusercontent.com/luckymehra/epidem-exercises/master/data/ex4.csv). Import the data and create new variables using the code below.

```{r data-import, eval=TRUE}

# import data
a <- read_csv("https://raw.githubusercontent.com/luckymehra/epidem-exercises/master/data/ex4.csv",
              col_types = cols(
                blk = col_factor(), # parse blk as a factor
                trt = col_factor() # parse trt as a factor
                ))

# create new variables
a$y <- a$pctsev/100
a$ystar <- log(a$y/(1-a$y))
a$wt <- a$y*(1-a$y)

# print the data
knitr::kable(head(a),
             caption = "The first 6 rows of dataset *a*.")

# get a glimpse of data
glimpse(a)
```


## First mixed model

### Fit the model

Run the mixed model analysis using **nlme** package in R. The function used to fit the mixed model is called `lme()`.

```{r first-mixed-model, eval = TRUE}

# fit the model

mm_1 <- lme(ystar ~ trt*t, # fixed effects
            data = a, 
            random = list(blk = ~ 1, plot = ~ 1), # random effects
            correlation = corAR1(form = (plot = ~ 1)), # specify that observations within a plot are autocorrelated. Other valid arguments are corAR1, corARMA, corCAR1, corCompSymm, corExp, corGaus, corLin, corRatio, corSpher, corSymm
            contrasts = list(trt = "contr.SAS"), # specify this option to get parameter estimates similar to reported by SAS
            weights = ~ I(1/wt))

# output the summary
summary(mm_1)

# extract covariance parameter estimates
VarCorr(mm_1)

# extract type3 fixed effects anova
anova.lme(mm_1, type = 'marginal')
```

### Diagnostic plots

```{r diagnostic-plots, eval=TRUE}

# pearson residuals vs. fitted values
plot(mm_1, resid(., type="pearson") ~ fitted(.), abline = 0)

# standardaized residuals vs. fitted values
plot(mm_1, resid(., scaled=TRUE) ~ fitted(.), abline = 0)

# qq plot
qqnorm(residuals(mm_1))
qqline(residuals(mm_1))

#observed vs. fitted values
plot(mm_1, ystar ~ fitted(.), abline = c(0,1))
```

## Second mixed model

### Fit the model

Run the mixed model analysis using **nlme** package in R. The function used to fit the mixed model is called `lme()`. Here we will specify no intercept. We will also use **emmeans** package to get least squared means and contrasts.

```{r second-mixed-model, eval = TRUE}

# fit the model
#library(nlme)
mm_2 <- update(mm_1, fixed = ystar ~ - 1 + trt + trt:t) # update fixed effects in mm_1, -1 indicates no intercept

# output the summary
summary(mm_2)

# extract covariance parameter estimates
VarCorr(mm_2)

# extract type3 fixed effects anova
anova.lme(mm_2, type = 'marginal')

# compare the slopes for different treatments
#library(emmeans)

emtrends(mm_2, pairwise ~ trt, var="t", adjust = "none")

# get the treatment difference at various time points
emmeans(mm_2, pairwise ~ trt|t, nesting = NULL, at = list(t = c(0, 7, 14, 21, 28, 35)), adjust = "none")
```

### Plot observed versus predicted model values

```{r obs-pred-plot, eval = TRUE}

# add fitted and residuals in to a new dataset called b
b = cbind(a, resid = resid(mm_2), fitted = fitted(mm_2))

# fit linear regression
b.lm <- lm(ystar ~ fitted, data=b)

# plot using ggplot2 package
ggplot(b, aes(x=fitted, y = ystar)) +
  geom_point(color="blue", size = 3) +
  geom_smooth(method = lm, color = "lightgrey")
```

