# Exercise 9.5 {#ex9_5}

## Load packages

Here is the R code to download the required packages for this exercise.

```{r pac-load, eval = TRUE}
# install package manager 'pacman'
if (!require(pacman)){
  install.packages('pacman')
}

# load packages needed for this exercise
library(pacman)
p_load(tidyverse,
       lctools, # to calculate Moran's I
       spdep, # to calculate geary's c
       geoR, # to compute variogram
       gridExtra, # to stack plots
       gstat, automap, # packages for variogram model selection
       sp # need a function called `coordinates`
       )

```

## Data

This is equivalent to data step in SAS. Here, the data is imported from a file `data.csv` using the function `read_csv`. This function will download the file directly from [here](https://raw.githubusercontent.com/luckymehra/epidem-exercises/master/data/ex9_5.csv).

```{r data-import, eval=TRUE}
# Import data
a <- read_csv("https://raw.githubusercontent.com/luckymehra/epidem-exercises/master/data/ex9_5.csv")

# print the data
a
```

## Autocorrelation statistics

```{r autocorr-statistics, eval = TRUE}

# visualize the data
ggplot(data = a) +
  geom_point(mapping = aes(x = COL, y = ROW, size = YI, color = YI)) +
  ggtitle("Spatial Distribution of YI Observation") +
  theme(plot.title = element_text(hjust = 0.5))

# calculate Moran's I
Coords <- a %>% 
  dplyr::select(COL, ROW)

mI <- moransI(Coords, Bandwidth = 1, a$YI)

# print Moran's I table
moran.table <- tribble(
  ~`Moran's I`, ~`Expected I`, ~`Z randomization`, ~`P value randomization`,
  #------------/--------------/-------------------/------------------------
  mI$Morans.I, mI$Expected.I,  mI$z.randomization, mI$p.value.randomization
  )

moran.table

# create Moran's I scatter plot
l.moran <- l.moransI(Coords, Bandwidth = 1, a$YI)

# calculate geary's c
Coords_num <- coordinates(Coords)

# create an object of class 'nb' so that it can be used with function from packege `spdep`
Coords_nb <- knn2nb(knearneigh(Coords_num))

# create a 'listw' object for use in the function `geary.test`
coords_listw <- nb2listw(Coords_nb)

gearyC <- geary.test(a$YI, coords_listw, alternative = "two.sided")
gearyC

```

## First variogram  

We will use the package `geoR` to construct empricial variogram, and then draw them using package `ggplot2`.  

```{r first-variogram-model, eval=TRUE}
v1 <- variog(coords = Coords_num, data = a$YI, breaks = seq(0.5, 15.5),
             max.dist = 12)

# extract data from object v1 for plotting
v1_plot_data <- cbind(v1$u, v1$v, v1$n) %>% 
  as.data.frame() %>% 
  dplyr::rename(Distance = V1,
                Semivariance = V2,
                Pair_count = V3)

# in the table below, gamma is semivariance
v1_plot_data

# plot variogram
v1_plot_vario <- ggplot(data = v1_plot_data) +
  geom_point(mapping = aes(x = Distance, y = Semivariance)) +
  ggtitle("Empirical Semivariogram of YI") +
  theme(plot.title = element_text(hjust = 0.5))

# plot pair counts
v1_plot_pair_count <- ggplot(data = v1_plot_data) +
  geom_col(mapping = aes(x = Distance, y = Pair_count), width = 0.01, color = "blue")

# stack two plots
grid.arrange(v1_plot_vario, v1_plot_pair_count,
             ncol = 1, heights = c(3, 1))
```

## Second variogram  

Plot robust and classical variogram together.  

```{r second-variogram-model, eval = TRUE}
# fit robust variogram
v1_robust <- variog(coords = Coords_num, data = a$YI, breaks = seq(0.5, 15.5),
             max.dist = 12, estimator.type = "modulus")

# extract the data
v1_robust_data <- cbind(v1_robust$u, v1_robust$v, v1_robust$n) %>% 
  as.data.frame() %>% 
  dplyr::rename(Distance = V1,
                Semivariance = V2,
                Pair_count = V3)

# plot robust variogram
v1_robust_vario <- ggplot(data = v1_robust_data) +
  geom_point(mapping = aes(x = Distance, y = Semivariance)) +
  ggtitle("Empirical Semivariogram of YI - Robust estimation") +
  theme(plot.title = element_text(hjust = 0.5))

v1_robust_vario

# combine robust and classical variogram  
var_comb <- v1_robust_data %>% 
  
  # combine robust and classical variogram datasets
  dplyr::rename(Semivariance_robust = Semivariance) %>% 
  bind_cols(dplyr::select(v1_plot_data, Semivariance)) %>% 
  gather(key = "Semivariance_type", value = "Semivariance", -c(Distance, Pair_count)) %>% 
  
  # plot
  ggplot() +
  geom_point(mapping = aes(x = Distance, y = Semivariance, color = Semivariance_type)) +
  ggtitle("Empirical Semivariogram for YI") +
  theme(plot.title = element_text(hjust = 0.5))

var_comb
```  

## Variogram model selection 

We will use the package `gstat` and `automap` for variogram model selection

```{r model-selection, eval=TRUE}
# specify coordinates in the dataset
coordinates(a) = ~COL+ROW

# select the best model out of exponential, spherical, and gaussian  
autofitVariogram(YI ~ COL + ROW, a, model = c("Sph", "Exp", "Gau"), cutoff = 12)

# fit empirical variogram
v_emp <- variogram(YI ~ COL + ROW, data = a, cutoff = 12)
v_emp
plot(v_emp)

# fit exponential variogram
v_exp <- fit.variogram(v_emp, vgm("Exp"))

# fit spherical and gaussian
v_sph <- fit.variogram(v_emp, vgm("Sph"))
v_sph

v_gau <- fit.variogram(v_emp, vgm("Gau"))

# extract plotting data from fitted variograms
v_exp_line <- variogramLine(v_exp, maxdist = 12)
v_sph_line <- variogramLine(v_sph, maxdist = 12)
# v_gau_line <- variogramLine(v_gau, maxdist = 12)

# plot emprical and fitted variograms together  
# specify color for legends
legend_color <- c("Empirical" = "blue", "Exponential" = "blue",
                  "Spherical" = "orange")
ggplot(data = v_emp) +
  geom_point(mapping = aes(x = dist, y = gamma, fill = "Empirical"), color = "blue") +
  geom_line(data = v_exp_line, mapping = aes(x = dist, y = gamma, color = "Exponential")) +
  geom_line(data = v_sph_line, mapping = aes(x = dist, y = gamma, color = "Spherical")) +
  # geom_line(data = v_gau_line, mapping = aes(x = dist, y = gamma, color = "Gaussian")) +
  scale_color_manual(name = "", values = legend_color) +
  scale_fill_manual(name = "", values = legend_color) +
  labs(x = "Distance",
       y = "Semivariance")

```

