# Yield loss {#yieldloss}

## Load packages

Here is the R code to download the required packages for this exercise.

```{r pac-load, eval = TRUE}
# install package manager 'pacman'
if (!require(pacman)){
  install.packages('pacman')
}

# load packages needed for this exercise
pacman::p_load(tidyverse,
       nlme,
       emmeans,
       predictmeans
       )
```

## Data

This is equivalent to the data step in SAS. Here, the data is imported from a file `yield_loss.csv` using the function `read_csv`. This function will download the data file direclty from [here](https://raw.githubusercontent.com/luckymehra/epidem-exercises/master/data/yield_loss.csv).

```{r data-import, eval=TRUE}
# Import data
a <- read_csv("https://raw.githubusercontent.com/luckymehra/epidem-exercises/master/data/yield_loss.csv")

# print the data
a

# specify that FUNG, TRT, and BLK are factors
a$FUNG <- as.ordered(as.factor(a$FUNG))
a$TRT <- as.ordered(as.factor(a$TRT))
a$BLK <- as.ordered(as.factor(a$BLK))
```

## Mixed model for response variable DS 

```{r mixed-model-DS, eval=TRUE}
# fit the model  
mm_1 <- lme(DS ~ TRT*FUNG, # fixed effects
            data = a,
            random = ~1|BLK/TRT) # read mm_1 as mixed model 1

# summary output
summary(mm_1)

# type 3 tests of fixed effects
anova(mm_1)

# visualize interaction
emmip(mm_1, TRT ~ FUNG)

# to do anova for random effects, we need to compare mm_1 with a model that only has fixed effects,
# we can use `gls()` function in `nlme` to fit the fixed effects model
fixed_model <- gls(DS ~ TRT * FUNG,
									 data = a)

# test the random effects in the model
anova(mm_1, fixed_model)

# least square means
test(emmeans(mm_1, "TRT"))
test(emmeans(mm_1, "FUNG"))

# pairwise difference
test(emmeans(mm_1, pairwise ~ TRT), adjust = "none")
test(emmeans(mm_1, pairwise ~ FUNG))
test(emmeans(mm_1, pairwise ~ TRT*FUNG), adjust = "none")

```

### Diagnostic plots

```{r diagnostic-plots, eval = TRUE}

# pearson residuals vs. fitted values
plot(mm_1, resid(., type="pearson") ~ fitted(.), abline = 0)

# standardaized residuals vs. fitted values
plot(mm_1, resid(., scaled=TRUE) ~ fitted(.), abline = 0)

# qq plot
qqnorm(residuals(mm_1))
qqline(residuals(mm_1))

#observed vs. fitted values
plot(mm_1, DS ~ fitted(.), abline = c(0,1))
```

## Mixed model for response variable YIELD 

```{r mixed-model-yield, eval=TRUE}
# fit the model  
mm_2 <- lme(YIELD ~ TRT*FUNG, # fixed effects
            data = a,
            random = ~1|BLK/TRT) # read mm_2 as mixed model 2

# summary output
summary(mm_2)

# type 3 tests of fixed effects
anova(mm_2)

# visualize interaction
emmip(mm_2, TRT ~ FUNG)

# to do anova for random effects, we need to compare mm_1 with a model that only has fixed effects,
# we can use `gls()` function in `nlme` to fit the fixed effects model
fixed_model_YIELD <- gls(YIELD ~ TRT * FUNG,
									 data = a)

# test the random effects in the model
anova(mm_2, fixed_model_YIELD)

# least square means
test(emmeans(mm_2, "TRT"))
test(emmeans(mm_2, "FUNG"))

# pairwise difference
test(emmeans(mm_2, pairwise ~ TRT), adjust = "none")
test(emmeans(mm_2, pairwise ~ FUNG))
test(emmeans(mm_2, pairwise ~ TRT*FUNG), adjust = "none")

```

### Diagnostic plots

```{r diagnostic-plots-yield, eval = TRUE}

# pearson residuals vs. fitted values
plot(mm_2, resid(., type="pearson") ~ fitted(.), abline = 0)

# standardaized residuals vs. fitted values
plot(mm_2, resid(., scaled=TRUE) ~ fitted(.), abline = 0)

# qq plot
qqnorm(residuals(mm_2))
qqline(residuals(mm_2))

#observed vs. fitted values
plot(mm_2, YIELD ~ fitted(.), abline = c(0,1))
```

## Linear regression between YIELD and DS  

```{r lm-yield-ds, eval=TRUE}
# fit `lm` model
lm_1 <- lm(YIELD ~ DS, data = a) 
summary(lm_1)
anova(lm_1)

# diagnostic plots  
residplot(lm_1)

```

### Linear regression between RY1 and DS

```{r lm-ry1-ds, eval=TRUE}
b <- a %>% 
	mutate(RY1 = YIELD/399.23843)

# fit linear regression model
lm_2 <- lm(RY1 ~ DS, data = b)
summary(lm_2)
anova(lm_2)

# diagnostic plots  
residplot(lm_2)

```

### Transform dataset a

```{r data-transform, eval=TRUE}
a_yield <- a %>% 
	dplyr::select(BLK, TRT, YIELD) %>% 
	arrange(BLK, TRT, YIELD) %>% 
	group_by(BLK, TRT) %>% 
	summarise(RY2 = YIELD[1]/YIELD[2]) %>% 
	ungroup()


a_ds <- a %>% 
	dplyr::select(BLK, TRT, DS) %>% 
	arrange(BLK, TRT, DS) %>% 
	group_by(BLK, TRT) %>% 
	summarise(CDS = DS[2]) %>% 
	ungroup()

a_new <- a_yield %>% 
	inner_join(a_ds) %>% 
	ungroup() %>% 
	mutate(BLK = parse_factor(as.character(BLK)),
				 TRT = parse_factor(as.character(TRT)))

# print the data
a_new
```

## Mixed model for RY2

```{r mixed-model-RY2, eval=TRUE}
# fit the model  
mm_3 <- lme(RY2 ~ TRT, # fixed effects
            data = a_new,
            random = ~1|BLK) # read mm_3 as mixed model 3

# summary output
summary(mm_3)

# type 3 tests of fixed effects
anova(mm_3)

# to do anova for random effects, we need to compare mm_1 with a model that only has fixed effects,
# we can use `gls()` function in `nlme` to fit the fixed effects model
fixed_model_RY2 <- gls(RY2 ~ TRT,
									 data = a_new)

# test the random effects in the model
anova(mm_3, fixed_model_RY2)

# pairwise difference
test(emmeans(mm_3, pairwise ~ TRT), adjust = "none")

# diagnostic plots  
residplot(mm_3)
```

## Linear regression between RY2 and CDS

```{r lm_ry2-cds, eval=TRUE}
# fit linear regression model
lm_3 <- lm(RY2 ~ CDS, data = a_new)
summary(lm_3)
anova(lm_3)

# diagnostic plots  
residplot(lm_3)
```

